MSstatsPTM statistical relative quantification of post-translational modifications in global proteomics experiments

### 2021-11-15 main/MSstatsPTM_Main.tex
Updated per author comments

This section describes the datasets and experiments used to evaluate the proposed method. The datasets were chosen to provide a performance benchmark of the proposed approach against existing methods when ground truth is known, as well as represent a variety of experimental designs and acquisition methods where the approach can be applied. To benchmark the dataset, computer simulations and a spike-in experiment were used. The computer simulations vary in experimental realism; the first simulation was perfectly clean, with many replicates and no missing values, while the second simulation introduced real world qualities, such as limited modified features and missing values. The spike-in experiment took the real world qualities a step further and allowed us to compare the methods in a real experiment where ground truth is known. Finally the proposed approach was used to model three biological experiments which differed in experimental design, acquisition strategy, and organism. These experiments were chosen to show the proposed method's applicability to many different types of experiments.

@@

In this simulation real world experimental conditions were introduced, including missing values and low number of modified features. The percentage of missing values and feature counts were determined by looking at the averages from the biological experiments in this paper. Modified peptides were simulated with 2 features, while unmodified proteins were still simulated with 10 features. Additionally, 20\% of observations for both modified and unmodified peptides were masked completely at random. Adding missing values and few representative modified features provides a more realistic expectation of model performance. Otherwise this simulation used the same parameters as in simulation 1, with 1000 modified peptides half of which were positives and the other half negative.

@@

Figure \ref{fig:data-structure} schematically illustrates a simplified version of the data structure resulting from a typical bottom-up experiment for quantitative analysis of PTMs, in which there are multiple layers of variation present. A PTM site is quantified with multiple spectral features, which vary in sequence (e.g., fully or partially cleaved peptides), ionization efficiency, charge states, etc. The number quantified features vary across replicate LC-MS/MS runs of the same sample, and across conditions. To perform adjustment with respect to protein abundance, features of unmodified peptides are used for the inference of underlying protein abundance. Typically, because of the enrichment step for PTMs, very few of those features are present in original LC-MS runs. For more accurate estimation of protein abundance, separate global proteomics data of unenriched samples are recommended. If unmodified features are unavailable for any given modified feature, unmodified intensity adjustment cannot be performed. As different levels of variability are present in the data, the log-intensities of the features for modified and unmodified peptides are modeled separately using two linear mixed models.

>>>>

This section describes the datasets and experiments used to evaluate the proposed method. The datasets were chosen to provide a performance benchmark of the proposed approach against existing methods when ground truth is known, as well as represent a variety of experimental designs and acquisition methods where the approach can be applied. To benchmark the approach, computer simulations and a spike-in experiment were used. The computer simulations vary in experimental realism; the first simulation was perfectly clean, with many replicates and no missing values, while the second simulation introduced real world qualities, such as limited modified features and missing values. The spike-in experiment took the real world qualities a step further and allowed us to compare the methods in a real experiment where ground truth is known. Finally the proposed approach was used to model three biological experiments which differed in experimental design, acquisition strategy, and organism. These experiments were chosen to show the proposed method's applicability to many different types of experiments.

@@

In this simulation real world experimental conditions were introduced, including missing values and low number of modified features. The percentage of missing values and feature counts were determined by looking at the averages from the biological experiments in this paper. Modified peptides were simulated with 2 features, while unmodified proteins were still simulated with 10 features. Additionally, 20\% of observations for both modified and unmodified peptides were missing completely at random. Adding missing values and few representative modified features provides a more realistic expectation of model performance. Otherwise this simulation used the same parameters as in simulation 1, with 1000 modified peptides half of which were positives and the other half negative.

@@

Figure \ref{fig:data-structure} schematically illustrates a simplified version of the data structure resulting from a typical bottom-up experiment for quantitative analysis of PTMs, in which there are multiple layers of variation present. A PTM site is quantified with multiple spectral features, which vary in sequence (e.g., fully or partially cleaved peptides), ionization efficiency, charge states, etc. The number quantified features vary across replicate LC-MS/MS runs of the same sample, and across conditions. To perform adjustment with respect to protein abundance, features of unmodified peptides are used for the inference of underlying protein abundance. Typically, because of the enrichment step for PTMs, very few of those features are present in original LC-MS runs. For more accurate estimation of protein abundance, separate global proteomics data of unenriched samples are recommended. If unmodified peptides are unavailable for any given modification, unmodified intensity adjustment cannot be performed. As different levels of variability are present in the data, the log-intensities of the features for modified and unmodified peptides are modeled separately using two linear mixed models.
